Categorical Foundations of E-Graphs With Binding

### 2025-07-10 POPL2026/paper/CategoricalFoundationsOfEGraphsWithBinding.tex
mass format
lots of polish

\documentclass[acmsmall,screen, nonacm, review, anonymous, authoryear]{acmart}

@@

\newcommand\HypI[1]{\textbf{HypI}(#1)}
\newcommand\MdaCospans{\textbf{MHypI}(\Sigma)}
\newcommand{\Ecospans}{{\catname{EHypI(\Sigma)}}}
\newcommand{\MdaEcospans}{{\catname{MEHypI(\Sigma)}}}
\newcommand{\WellTypedMdaEcospans}{{\catname{MEHypI(\Sigma)}}}
% conflict
\newcommand{\hashtag}{{\#}}
\newcommand{\consistency}{{\smile}}

@@

\section{Conclusion}
In this work we introduced e-graphs with bindings formalised as morphisms of a closed $\catname{SLat}$-SMC.
We then provided a combinatorial representation for such morphisms in terms of hierarchical hypergraphs (e-hypergraphs) with a suitable notion of sound and complete DPO rewriting, making it possible for an e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
An important future work is exploring whether such combinatorial structures may lead to better algorithmics for doing equality saturation for $\lambda$-calculus.

\bibliographystyle{acm}
\bibliography{../../bibliography}

>>>>

\documentclass[acmsmall,screen, nonacm, review, anonymous]{acmart}

@@

\newcommand\HypI[1]{\textbf{HypI}(#1)}
\newcommand\MdaCospans{\textbf{MHypI}(\Sigma)}
\newcommand{\Ecospans}{{\catname{CEHypI(\Sigma)}}}
\newcommand{\MdaEcospans}{{\catname{MCEHypI(\Sigma)}}}
\newcommand{\WellTypedMdaEcospans}{{\catname{MCEHypI(\Sigma)}}}
% conflict
\newcommand{\hashtag}{{\#}}
\newcommand{\consistency}{{\smile}}

@@

\section{Conclusion}
In this work we introduced e-graphs with bindings formalised as morphisms of a closed symmetric monoidal $\catname{SLat}$-category.
We then provided a combinatorial representation for such morphisms in terms of hierarchical hypergraphs (e-hypergraphs) with a suitable notion of sound and complete DPO rewriting, making it possible for an e-hypergraph to represent equivalences derivable from a set of equations $\mathcal{E}$.
An important future work is exploring whether such combinatorial structures may lead to better algorithmics for doing equality saturation for $\lambda$-calculus.

\bibliographystyle{ACM-Reference-Format}
\bibliography{../../bibliography}

### 2025-07-10 POPL2026/paper/sections/appendix.tex
lots of polish

    \begin{proposition}[Special case of Proposition 6.4.6~\cite{Borceux_1994}]
      The forgetful functor $U : \catname{SLat} \to \catname{Set}$ has a left adjoint free functor $F : \catname{Set} \to \catname{SLat}$.
    \end{proposition}
    \begin{proof}
      The functor $F$ is defined by letting $F(A) = \coprod_{A} I_{\catname{SLat}}$ where the latter is a coproduct which is a `free' product of semilattices defined as follows.
      The elements of $S_{1} \coprod S_{2}$ are sequences $s_{1} + s_{2} + \ldots + s_{n}$ where each $s_{i}$ is either from $S_{1}$ or $S_{2}$ quotiented by all relations in $S_{1}$ and $S_{2}$.
      The adjunction is then given by the following natural isomorphisms
      \begin{align*}
      \catname{SLat}(\coprod_{A}(I_{\catname{SLat}}), B) &\cong \prod_{A}(\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A,\catname{SLat}(I_{\catname{SLat}}, B))\\
                                                         &\cong \catname{Set}(A, U(B))
      \end{align*}
      The fist isomorphism is given by the fact that hom-functor makes limits into colimits in its first argument, the second being given by the property that $|\catname{Set}(A,B)| = |B|^{|A|} = |\underbrace{B \times \ldots \times B}_{|A|}|$, and the last isomorphism is given by the definition of $U$.
      Furthermore, we have, 
      \[
      F(I) \cong I_{\catname{SLat}}
      \]
      and 
      \begin{align*}
      F(A) \otimes F(B) &\cong (\coprod_{A} I_{\catname{SLat}}) \otimes (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (I_{\catname{SLat}} \otimes \coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A} (\coprod_{B} (I_{\catname{SLat}} \otimes I_{\catname{SLat}}))\\
            &\cong \coprod_{A} (\coprod_{B} I_{\catname{SLat}})\\
            &\cong \coprod_{A \times B} I_{\catname{SLat}}\\
            &\cong F(A \times B)
      \end{align*}
    In the above we used the fact that functors $- \otimes X$ and $X \otimes -$ preserve colimits as they are both left-adjoint by symmetry and monoidal closedness of $\catname{SLat}$.
    \end{proof}

@@

We next explicitly construct a pushout.
Consider the diagram below.
    \[
        \adjustbox{scale=1.25}{
            \begin{tikzcd}
            Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
            Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
                                                                            &                                                                  & Q
            \end{tikzcd}}
    \]
    Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
    First, a coproduct of $X+Y$ is computed which is 
    \[
        X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
    \]
    where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
    We will omit labels as they are irrelevant to pushout construction.
    Consider relations
    \begin{align*}    
    S_{V} &= \{
          \;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
          \}\\
          &\cup\\
          &\{
          \;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; | \\
          &\;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z)\\
          &\;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
          \}\\
          &\cup\\
          &\{
          \;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}
          \}\\
        \\
    S_{E} &= \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
    \end{align*}
and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

\begin{align*}
        \sfrac{X + Y}{\sim (R_{V},R_{E})} = &(
            \sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},\\
            &\sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},\\
            &\consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},\\
            &\textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}\\
            &\textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}})    
\end{align*}

@@

    \begin{enumerate}
        \item If there exists $v$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$ and $u \sim v$, we let
              \ifdefined \ONECOLUMN
              \[
                \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))
                =
                [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)))
              \]
              \else
              \begin{align*}
                &\consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))\\
                &=\\
                &[[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)))
            \end{align*}
            \fi
        \item $u$ has no pre-image in $V_{Z}$ and $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(u)) = \varnothing$ and there exists $v$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$ and such that there is an undirected path from $[u]$ to $[v]$.
        Then we let
        \[
            \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u])) = \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))
        \]
    \end{enumerate}

@@

    This construction is analogous to how a pushout was constructed in~\cite{tiurin2025equivalencehypergraphsdporewriting} and we refer the reader for more detailed proofs there.
    The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in Fig.~\ref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
    \adjustbox{scale=0.8}{
	\begin{pgfonlayer}{nodelayer}
		\node [style=node, label={above:$v_1$}] (0) at (-5, 3.5) {};
		\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
		\node [style=node, label={above:$v_3$}] (2) at (-4, 3.5) {};
		\node [style=node, label={above:$v'_1$}] (3) at (-3.5, 3.5) {};
		\node [style=node, label={above:$v'_2$}] (4) at (-3, 3.5) {};
		\node [style=node, label={above:$v'_3$}] (5) at (-2.5, 3.5) {};
	\end{pgfonlayer}
    }
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of Theorem~\ref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ are given in \autoref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

\section{DPOI rewriting}
\label{sec:appendix:dpoi}

Consider an example of DPOI rewriting in \autoref{fig:dpoi-example} as defined in Definition~\ref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of \autoref{def:dpoi-e}.

\begin{figure}
    \ifdefined\ONECOLUMN
    \[
    \adjustbox{width=0.6\linewidth}{
    \tikzfig{./figures/dpo_example}
    }
    \]
    \else
    \[
\adjustbox{width=\linewidth}{
\tikzfig{./figures/dpo_example}
}
\]
    \fi
    \caption{$\beta$-rule DPOI example.}
    \label{fig:dpoi-example}
\end{figure}

>>>>

\begin{proposition}[Special case of Proposition 6.4.6~\cite{Borceux_1994}]
	The forgetful functor $U : \catname{SLat} \to \catname{Set}$ has a left adjoint free functor $F : \catname{Set} \to \catname{SLat}$.
\end{proposition}
\begin{proof}
	The functor $F$ is defined by letting $F(A) = \coprod_{A} I_{\catname{SLat}}$ where the latter is a coproduct which is a `free' product of semilattices defined as follows.
	The elements of $S_{1} \coprod S_{2}$ are sequences $s_{1} + s_{2} + \ldots + s_{n}$ where each $s_{i}$ is either from $S_{1}$ or $S_{2}$ quotiented by all relations in $S_{1}$ and $S_{2}$.
	The adjunction is then given by the following natural isomorphisms
	\begin{align*}
		\catname{SLat}(\coprod_{A}(I_{\catname{SLat}}), B) & \cong \prod_{A}(\catname{SLat}(I_{\catname{SLat}}, B))       \\
		                                                   & \cong \catname{Set}(A,\catname{SLat}(I_{\catname{SLat}}, B)) \\
		                                                   & \cong \catname{Set}(A, U(B))
	\end{align*}
	The fist isomorphism is given by the fact that hom-functor makes limits into colimits in its first argument, the second being given by the property that $|\catname{Set}(A,B)| = |B|^{|A|} = |\underbrace{B \times \ldots \times B}_{|A|}|$, and the last isomorphism is given by the definition of $U$.
	Furthermore, we have,
	\[
		F(I) \cong I_{\catname{SLat}}
	\]
	and
	\begin{align*}
		F(A) \otimes F(B) & \cong (\coprod_{A} I_{\catname{SLat}}) \otimes (\coprod_{B} I_{\catname{SLat}}) \\
		                  & \cong \coprod_{A} (I_{\catname{SLat}} \otimes \coprod_{B} I_{\catname{SLat}})   \\
		                  & \cong \coprod_{A} (\coprod_{B} (I_{\catname{SLat}} \otimes I_{\catname{SLat}})) \\
		                  & \cong \coprod_{A} (\coprod_{B} I_{\catname{SLat}})                              \\
		                  & \cong \coprod_{A \times B} I_{\catname{SLat}}                                   \\
		                  & \cong F(A \times B)
	\end{align*}
	In the above we used the fact that functors $- \otimes X$ and $X \otimes -$ preserve colimits as they are both left-adjoint by symmetry and monoidal closedness of $\catname{SLat}$.
\end{proof}

@@

	We next explicitly construct a pushout.
	Consider the diagram below.
	\[
		\adjustbox{scale=1.25}{
			\begin{tikzcd}
				Z \arrow[r, "f"] \arrow[d, "g"']                                   & X \arrow[d, "\sfrac{\iota_1}{\sim R}"] \arrow[rdd, "j_1", bend left] &   \\
				Y \arrow[r, "\sfrac{\iota_2}{\sim R}"] \arrow[rrd, "j_2"', bend right] & \sfrac{X+Y}{\sim R} \arrow[rd, "u"]                              &   \\
				&                                                                  & Q
			\end{tikzcd}}
	\]
	Then, the pushout of e-hypergraphs $X$ and $Y$ is computed in two steps.
	First, a coproduct of $X+Y$ is computed which is
	\[
		X + Y = \{V_{X} + V_{Y}, E_{X} + E_{Y}, s_{X+Y}, t_{X+Y}, \consistency_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}\}
	\]
	where $s_{X+Y} : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ which can be defined as a copairing $[s'_{X}, s'_{Y}] : E_{X} + E_{Y} \to (V_{X} + V_{Y})^{*}$ and $s'_{X} : E_{X} \to (V_{X} + V_{Y})^{*}$, $s'_{Y} : E_{Y} \to (V_{X} + V_{Y})^{*}$ defined as $s'_{X} = s_{X};\iota_{1,V}^{*}$ and $s'_{Y} = s_{Y};\iota_{2,V}^{*}$ where $\iota_1,\iota_2$ are corresponding coproduct injections, similarly for $t_{X+Y}, \textcolor{e-color}{<_{X+Y}}, \textcolor{closed-color}{<_{X+Y}}, \consistency_{X+Y}$.
	We will omit labels as they are irrelevant to pushout construction.
	Consider relations
	\begin{align*}
		S_{V} & = \{
		\;(x_i,y_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(y_j,x_j) \in (V_{X} + V_{Y}) \times (V_{X} + V_{Y})\; |                                               \\
		      & \;\exists z \in V_{Z} \; . \; x_i = f_{V};\iota_{1,V}(z) \text{ and } y_j = g_{V};\iota_{2,V}(z) \\
		      & \;\text{ where $x_i \in V_{X}$ and $y_j \in V_{Y}$ }
		\}                                                                                                       \\
		      & \cup                                                                                             \\
		      & \{
		\;(x,x)\;\text{where}\; x \in V_{X} + V_{Y}
		\}                                                                                                       \\
		\\
		S_{E} & = \{(x,x) \text{ where } x \in E_{X} + E_{Y}\}
	\end{align*}
	and let relations $R_{V},R_{E}$ be their transitive closures respectively.

@@

	\begin{align*}
		\sfrac{X + Y}{\sim (R_{V},R_{E})} = & (
		\sfrac{V_{X} + V_{Y}}{\sim (R_{V},R_{E})},                                                             \\
		                                    & \sfrac{E_{X} + E_{Y}}{\sim (R_{V},R_{E})},                       \\
		                                    & \sfrac{s_{X+Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \sfrac{t_{X,Y}}{\sim (R_{V},R_{E})},                             \\
		                                    & \consistency_{\sfrac{X+Y}{\sim (R_{V},R_{E})}},                  \\
		                                    & \textcolor{e-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}}       \\
		                                    & \textcolor{closed-color}{<}_{\sfrac{X + Y}{\sim (R_{V},R_{E})}})
	\end{align*}

@@

	\begin{enumerate}
		\item If there exists $v$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$ and $u \sim v$, we let
		      \ifdefined \ONECOLUMN
			      \[
				      \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))
				      =
				      [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)))
			      \]
		      \else
			      \begin{align*}
				       & \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u]))           \\
				       & =                                                                                               \\
				       & [[]_{V}^{\consistency},[]_{E}^{\consistency}]^{*}(\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)))
			      \end{align*}
		      \fi
		\item $u$ has no pre-image in $V_{Z}$ and $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(u)) = \varnothing$ and there exists $v$ such that $\consistency_{X+Y}(\iota_{V_{X} + V_{Y}}(v)) \not = \varnothing$ and such that there is an undirected path from $[u]$ to $[v]$.
		      Then we let
		      \[
			      \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([u])) = \consistency_{\sfrac{X+Y}{\sim}}^{\hashtag}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}([v]))
		      \]
	\end{enumerate}

@@

	This construction is analogous to how a pushout was constructed by~\citet{tiurin2025equivalencehypergraphsdporewriting} and we refer the reader for more detailed proofs there.
	The only addition is $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$ that is constructed analogously to $\textcolor{e-color}{<_{\sfrac{X+Y}{\sim}}^{\mu}}$.
\end{proof}
The construction can be illustrated with the example in~\autoref{fig:pushout_example}.
Vertices and edges that are being identified as they have a common preimage inherit the relations from the vertex or edge that has it defined either directly or via a path.
In particular, a coproduct of two feet contains\begin{tikzpicture}
	\adjustbox{scale=0.8}{
		\begin{pgfonlayer}{nodelayer}
			\node [style=node, label={above:$v_1$}] (0) at (-5, 3.5) {};
			\node [style=node, label={above:$v_2$}] (1) at (-4.5, 3.5) {};
			\node [style=node, label={above:$v_3$}] (2) at (-4, 3.5) {};
			\node [style=node, label={above:$v'_1$}] (3) at (-3.5, 3.5) {};
			\node [style=node, label={above:$v'_2$}] (4) at (-3, 3.5) {};
			\node [style=node, label={above:$v'_3$}] (5) at (-2.5, 3.5) {};
		\end{pgfonlayer}
	}
\end{tikzpicture}
such that $\textcolor{closed-color}{<_{X+Y}}$ is only defined for $v_1,v_2,v_3$.
Therefore, we assign $\textcolor{closed-color}{<_{\sfrac{X+Y}{\sim}}}(\iota_{\sfrac{V_{X} + V_{Y}}{\sim}}[v_1']) = [\textcolor{closed-color}{<_{X+Y}}(\iota_{V_{X} + V_{Y}}(v_1))]$ and so on.
This motivates the requirement for conditions (2), (3), and (4) of~\autoref{th:existence_of_pushouts}, otherwise such inheritance would be ambiguous.

@@

The base cases for interpretation function $[-] : \textbf{SMT}(\Sigma) \to \MdaCospans$ are given in~\autoref{fig:base_cases}.
Then it is defined inductively by letting $[f \otimes g] \Coloneqq [f] \otimes [g]$ and $[f;g] \Coloneqq [f];[g]$ and by freeness induces the corresponding functor.

\section{EDPOI rewriting}%
\label{sec:appendix:dpoi}

Consider an example of EDPOI rewriting in~\autoref{fig:dpoi-example} as defined in~\autoref{def:dpoi-e}.
The rule in the top half of the diagram is an instance of a $\beta$-reduction rule as defined for string diagrams~\cite{ghica2024stringdiagramslambdacalculifunctional}.
Note that the match satisfies all the conditions of the boundary complement as all vertices from the interface have the same parent.
First, to compute the boundary complement, we remove from the source e-hypergraph everything that has no pre-image in the interface above.
The mda-cospan for the complement then will be
\[
	[u_1] \to [u_1, v_1, v_2] + [v_3] \to \mathcal{L}^{\bot} \xleftarrow{} [v_3] + [v_1,v_2] \xleftarrow{} [u_2] ~.
\]
Note how we removed strict internal interfaces of $\mathcal{L}$ from internal interfaces of $\mathcal{G}$, yielding $[u_1,v_1,v_2]$ and similarly for output interfaces.
After computing the pushout for the second half of the diagram we add strict internal interfaces of $\mathcal{R}$ but since there are none, we are done as per condition (2) of~\autoref{def:dpoi-e}.

\begin{figure}
	\ifdefined\ONECOLUMN
		\[
			\adjustbox{width=0.6\linewidth}{
				\tikzfig{./figures/dpo_example}
			}
		\]
	\else
		\[
			\adjustbox{width=\linewidth}{
				\tikzfig{./figures/dpo_example}
			}
		\]
	\fi
	\caption{$\beta$-rule EDPOI example.}
	\label{fig:dpoi-example}
\end{figure}

### 2025-07-10 POPL2026/paper/sections/applications.tex
lots of polish

While the work of~\cite{slotted-egraphs} has made the variables and binders into first class citizens of an e-graph, substitution and $\beta$-reduction in the context of $\lambda$-calculus remain challenging.
The issue with substitution is that despite it acting on a small part of the term, it results in duplication of large portions of an e-graph which is practically undesirable.
While it is possible to define substitution as a built-in operation in a slotted e-graph, in practice, using explicit substitution is more appropriate as it allows for a finer grained control over the duplication.
Even more so, explicit substitution becomes mandatory if we want to encode lambda calculus with explicit substitution in an e-graph.
Below is the Table~\ref{tbl:slotted} that features the number of e-nodes when performing equality saturation for a $\beta$-reduction using a lambda term $plus 2 2$ using Church encoding of arithmetic.

\begin{table}
    \begin{tabular}{lcccc}
      Set of equations & built-in subst & explicit subst & explicit subst & explicit subst\\
      &&& + interchange&  + interchange\\
      &&&&  + comm\\
      \# of e-nodes & 40 & 161 & 289 & 356
    \end{tabular}
    \caption{Slotted e-graphs benchmarks}
    \label{tbl:slotted}
\end{table}

>>>>

While the work of~\cite{slotted-egraphs} has made the variables and binders into first class citizens of an e-graph, substitution and $\beta$-reduction in the context of $\lambda$-calculus remain challenging.
The issue with substitution is that despite it acting on a small part of the term, it results in duplication of large portions of an e-graph which is practically undesirable.
While it is possible to define substitution as a built-in operation in a slotted e-graph, in practice, using explicit substitution is more appropriate as it allows for a finer grained control over the duplication.
Even more so, explicit substitution becomes mandatory if we want to encode lambda calculus with explicit substitution in an e-graph.
Below is the~\autoref{tbl:slotted} that features the number of e-nodes when performing equality saturation for a $\beta$-reduction using a lambda term $plus 2 2$ using Church encoding of arithmetic.

\begin{table}
	\begin{tabular}{lcccc}
		Set of equations & built-in subst & explicit subst & explicit subst & explicit subst \\
		                 &                &                & + interchange  & + interchange  \\
		                 &                &                &                & + comm         \\
		\# of e-nodes    & 40             & 161            & 289            & 356
	\end{tabular}
	\caption{Slotted e-graphs benchmarks}
	\label{tbl:slotted}
\end{table}
